---
title: "Introduction to R"
author: "Enrique Valencia Lopez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
---

# Introduction to R 

Welcome to this introductory R  notebook! This tutorial will guide you through the basics of R. 
Our goals for today's workshop are:

1.  Demonstrate how to navigate R-Studio.

2.  Define what R is useful and not useful for.

3.  Use code to create variables and vectors.

4. Some basic functions and (very basic) plots

Follow along by running each code chunk and trying the exercises.

# Preliminary: The Global Environment

```{r}
###Remove objects in the global environment 
cat("\014") #Clear console
rm(list = ls())  # Remove all objects in the global environment

```

All variables created show up in the Global Environment (top right panel).

```{r}
# print a list of all variables in your environment
ls()
```
Let's create some 
```{r}
year<-c(1998, 1999, 2000)
flowers<-c("Rose", "Tulips", "Marygold")
famous_cats<-c("Garfield", "Puss in Boots", "Snowbell", "Cheshire Cat")

```

You can remove variables from the global environment.

```{r}

# remove a single variable
rm(famous_cats)

```

```{r}
# remove all variables in the global environment
rm(list = ls())
```



## 1. Variables and Data Types

In R, we can store different types of data in variables. Let's explore the main data types:

```{r}
# Create numeric, character, and logical variables
num_variable <- 42
char_variable <- "Hello, World"
logic_variable <- TRUE

# Print the variables
print(num_variable)
print(char_variable)
print(logic_variable)

# Check their data types
class(num_variable)
class(char_variable)
class(logic_variable)
```

### 💫 Your Turn!
Create three new variables of different types and print their classes:
- A numeric variable called `Year_of_Birth`
- A character variable called `My_Fav_Sport`
- A logical variable called `Need_Break`

```{r}
# Write your code here


```

## 2. Vectors

Vectors are one-dimensional arrays that can hold data of the same type. They're one of the most fundamental data structures in R.

```{r}
# Create different types of vectors
num_vector <- c(2024, 2025, 2026, 2027)
char_vector <- c("Joe Bruin", "Peter the Anteater", "Oski Bear")
logic_vector <- c(TRUE, FALSE, TRUE)
mixed_vector<-c(1,2,3, "Tree") ##What class do you think this vector will be

# Print the vectors
print(num_vector)
print(char_vector)
print(logic_vector)
print(mixed_vector)

# Access elements of a vector (R uses 1-based indexing)
print(num_vector[3])  # Get the third element
print(char_vector[2]) # Get the second element
```

### 💫 Your Turn!
Create a vector of your three activities you enjoy doing in your free time and then access the second activity in your vector.

```{r}
# Write your code here
```

## 3. Basic Operations

R can perform various mathematical operations. 

```{r}
# Basic arithmetic
a <- 10
b <- 3

# Addition
print(a + b)

# Subtraction
print(a - b)

# Multiplication
print(a * b)

# Division
print(a / b)

# Integer division
print(a %/% b)

# Modulus (remainder)
print(a %% b)

# Powers
print(a ^ 2)

#You can also do:
  print(1+2)
```

### 💫 Your Turn!
Calculate the area and perimeter of a rectangle with length 8 and width 5 inches.

```{r}
# Write your code here
```

## 4. Functions

R has many built-in functions that help us analyze data:

```{r}
# Create a vector of random numbers
random_vector <- sample(1:100, 20)

# Calculate basic statistics
print(mean(random_vector))    # Average
print(median(random_vector))  # Median
print(sd(random_vector))      # Standard deviation
print(sum(random_vector))     # Sum
print(length(random_vector))  # Length

# Sort the vector
sorted_vector <- sort(random_vector)
print(sorted_vector)
```

## 5. Data Frames

Data frames are table-like structures that can hold different types of data:

```{r}
# Create a data frame
UCs <- data.frame(
  name = c("Peter the Anteater", "Joe the Bruin", "Oski Bear", "King Triton"),
  age = c(59, 105, 156, 63),
  length_campus = c(1475, 419, 178, 1976),
  Southern_California = c(TRUE, TRUE, FALSE, TRUE)
)

# View the data frame
print(UCs)

# Access columns
print(UCs$name)
print(UCs$age)

# Get basic information about the data frame
str(UCs)
summary(UCs)
```

## 6. Basic Plotting

R has some powerful plotting capabilities. 

```{r}
# Create a histogram
hist(UCs$age, 
     main = "Distribution of Ages of the UCs",
     xlab = "Age",
     ylab = "Frequency",
     col = "lightblue",
     border = "white")

# Create a scatter plot
plot(UCs$age, UCs$length_campus,
     main = "Age of the UCs vs. Length of Campus",
     xlab = "Age of the UCs",
     ylab = "Length of campus",
     pch = 19, 
     col = "blue")
```

For  pch, see [plotting symbols](http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r)

## 7. Final Exercise

Now let's put everything together! Complete the following tasks:

1. Create a data frame containing information about 5 movies you enjoy:
   - Include columns for: title, year, your personal rating (1-10), and genre
2. Calculate the average rating of all five movies
3. Create a histogram showing the ratings for each movie
4. Find the oldest and newest movies in your dataset

```{r}
# Write your solution here
```

## Congratulations! 🎉
You've completed part 1.
Part of this workshop was based on [Berkeley's R Fundamental's](https://github.com/dlab-berkeley/R-Fundamentals) and [UC Irvine's R online resources](https://guides.lib.uci.edu/dataanalysis/r)

