# Merging Data

In this notebook we will explore how to merge data using SQL Joins, but in R. 
SQL joins are the foundation of database management systems, enabling the combination of data from multiple tables based on relationships between columns. 



## Setup

First, let's create our sample datasets and load required libraries:

```{r}
library(dplyr)

# Create sample datasets
customers <- data.frame(
  student_id = c(1, 2, 3, 4),
  name = c("Oski", "Peter", "Joe", "Tree"),
  stringsAsFactors = FALSE
)

orders <- data.frame(
  order_id = c(101, 102, 103, 104),
  customer_id = c(1, 2, 2, 5),
  amount = c(100, 150, 200, 300)
)
```

```{r}
customers
```


```{r}
orders

```

## 1. Left Join

<svg width="200" height="150" viewBox="0 0 200 150">
  <!-- Left circle (customers) -->
  <circle cx="85" cy="75" r="50" fill="#7cb9e8" fill-opacity="0.6" stroke="black"/>
  <!-- Right circle (orders) -->
  <circle cx="115" cy="75" r="50" fill="none" stroke="black" stroke-dasharray="5,5"/>
  <!-- Labels -->
  <text x="45" y="75" font-size="12">Customers</text>
  <text x="135" y="75" font-size="12">Orders</text>
</svg>

Left join keeps all records from the left table (customers) and matching records from the right table (orders). Non-matching records from the left table are filled with NA.

```{r}
left_result <- left_join(customers, orders, by = "customer_id")
print(left_result)
```

## 2. Right Join

<svg width="200" height="150" viewBox="0 0 200 150">
  <!-- Left circle (customers) -->
  <circle cx="85" cy="75" r="50" fill="none" stroke="black" stroke-dasharray="5,5"/>
  <!-- Right circle (orders) -->
  <circle cx="115" cy="75" r="50" fill="#7cb9e8" fill-opacity="0.6" stroke="black"/>
  <!-- Labels -->
  <text x="45" y="75" font-size="12">Customers</text>
  <text x="135" y="75" font-size="12">Orders</text>
</svg>

Right join keeps all records from the right table (orders) and matching records from the left table (customers). Non-matching records from the right table are filled with NA.

```{r}
right_result <- right_join(customers, orders, by = "customer_id")
print(right_result)
```

## 3. Inner Join

<svg width="200" height="150" viewBox="0 0 200 150">
  <!-- Left circle (customers) -->
  <circle cx="85" cy="75" r="50" fill="none" stroke="black"/>
  <!-- Right circle (orders) -->
  <circle cx="115" cy="75" r="50" fill="none" stroke="black"/>
  <!-- Intersection -->
  <path d="M 115,75 m -30,0 a 30,30 0 0,0 60,0 a 30,30 0 0,0 -60,0" 
        fill="#7cb9e8" fill-opacity="0.6"/>
  <!-- Labels -->
  <text x="45" y="75" font-size="12">Customers</text>
  <text x="135" y="75" font-size="12">Orders</text>
</svg>

Inner join only keeps records that exist in both tables. No NA values in the joining columns.

```{r}
inner_result <- inner_join(customers, orders, by = "customer_id")
print(inner_result)
```

## 4. Full Join

<svg width="200" height="150" viewBox="0 0 200 150">
  <!-- Left circle (customers) -->
  <circle cx="85" cy="75" r="50" fill="#7cb9e8" fill-opacity="0.6" stroke="black"/>
  <!-- Right circle (orders) -->
  <circle cx="115" cy="75" r="50" fill="#7cb9e8" fill-opacity="0.6" stroke="black"/>
  <!-- Labels -->
  <text x="45" y="75" font-size="12">Customers</text>
  <text x="135" y="75" font-size="12">Orders</text>
</svg>

Full join keeps all records from both tables. Non-matching records are filled with NA.

```{r}
full_result <- full_join(customers, orders, by = "customer_id")
print(full_result)
```

## 5. Anti Join

<svg width="200" height="150" viewBox="0 0 200 150">
  <!-- Left circle (customers) -->
  <circle cx="85" cy="75" r="50" fill="#7cb9e8" fill-opacity="0.6" stroke="black"/>
  <!-- Right circle (orders) -->
  <circle cx="115" cy="75" r="50" fill="none" stroke="black"/>
  <!-- Hide intersection -->
  <path d="M 115,75 m -30,0 a 30,30 0 0,0 60,0 a 30,30 0 0,0 -60,0" 
        fill="white"/>
  <!-- Labels -->
  <text x="45" y="75" font-size="12">Customers</text>
  <text x="135" y="75" font-size="12">Orders</text>
</svg>

Anti join returns only the records from the first table that have no match in the second table.

```{r}
anti_result <- anti_join(customers, orders, by = "customer_id")
print(anti_result)
```

## Key Points

1. The `by` parameter specifies which column(s) to join on
2. You can join on multiple columns using `by = c("col1", "col2")`
3. For different column names, use `by = c("customer_id" = "id")`
4. Join order matters (except in full join)
5. NA values are used for non-matching records in left, right, and full joins

